---
title: "Total exportations of Mexico by State and Activity"
author: Manuel T.
date: 2019-12-15
output:
  md_document:
    variant: markdown_github
---

# INTRODUCTION

The data set was taken from the website of the national institute of 
informatics. Downloaded in excel format, it contains info about annual 
exportation of products by sector, given in thousands of USD.

# IMPORTING AND TIDYING THE DATA 

The excel file is separated in different sheets per state. Each sheet 
contains the same information: annual exportation of products by sector, 
given in thousands of USD, where each column contains information of the
year, from 2007 to 2017.

Our first step will be to import the data from the excel file, and tidying
it in a useful way:

```{r, message = FALSE, warning = FALSE}

  ## Calling libraries
  library(readxl)
  library(tidyverse)

  ## Import from Excel

  ## The sheet names are the name of each state
  estados <- excel_sheets('ignore/EAEF_Entidad_Subsector.xlsx')
  ## Import column names and avoiding useless rows
  exportation <- read_xlsx('ignore/EAEF_Entidad_Subsector.xlsx',
                           range = "A5:N5", sheet = 1,
                           col_types = c("numeric", "text",
                                         rep("numeric", 12)))
  ## Create state column
  exportation <- mutate(exportation, state = "Aguascalientes")

  ## Iterate over every sheet
  for (i in 1:length(estados)){
      extracted <- read_xlsx('ignore/EAEF_Entidad_Subsector.xlsx',
                             range = "A7:N31", sheet = i,
                             col_names = F, na = "-",
                             col_types = c("numeric", "text",
                                           rep("numeric", 12)))
      colnames(extracted) <- colnames(exportation)[1:14]
      extracted <- mutate(extracted,
                          state = estados[i])
      exportation <- full_join(exportation, extracted)
  }

  ## Fix couple of column names
  colnames(exportation)[c(3, 14)] <- c("2007", "2018")
  exportation$`Código` <- parse_factor(as.character(exportation$`Código`))

  ## Tidying data

  export1 <- exportation %>%
      pivot_longer(cols = `2007`:`2018`,
                   names_to = "year",
                   values_to = "USD",
                   values_drop_na = T) %>%
      mutate(year = parse_double(year))
  export <- export1 %>%
      pivot_wider(names_from = "Descripción",
                  values_from = USD,
                  id_cols = c("state", "year"))

  ## Save our new data frames in a useful way
  write_csv(export1, "exportations_activity_rows.csv")
  write_csv(export, "exportations_activity_cols.csv")

```

Now we have 2 csv tables, `exportations_activity_rows.csv` (created from
`export1`) which keeps each activity as a row, and 
`exportations_activity_cols` (from `export`) which transform each activity
into a column using `pivot_wider`. Each of the 2 can help us to analyse
the data in different ways. We decided to keep them as csv files to make
its access easier for github users.

# Exploratory Data Analysis (EDA)

Now if we start a new session, all we have to do is to call our libraries
and load the data from CSV:

```{r, include = T, message = FALSE, warning = FALSE}
  library(tidyverse)

  export.rows <- read_csv("exportations_activity_rows.csv")
  export.cols <- read_csv("exportations_activity_cols.csv")
```

Due to the complexity in the variables names, we are going to storage them 
into an object to facilitate its accesibility. We could also rename them,
but this would reduce its descriptive nature.

```{r, include = T, message = FALSE, warning = FALSE}
  categorias <- colnames(export.cols)[3:27]
  categorias
```

Let's look at the totals by state, using `export.cols`

```{r, include = T, message = FALSE, warning = FALSE}
  export.cols %>%
      group_by(state) %>%
      summarise(`total export` = sum(!!sym(categorias[1]))) %>%
      arrange(desc(`total export`)) %>%
      print(n = Inf)

   ## Graphical mode
  export.cols %>%
      group_by(state) %>%
      summarise(`total export` = sum(!!sym(categorias[1]))) %>%
      ggplot() +
      geom_bar(aes(y = `total export`,
                   x = reorder(state, `total export`, FUN = abs),
                   fill = `total export`),
               stat = 'identity') +
      coord_flip()
```

Now, to do the same but by category, we could simply use `export.rows`

```{r, include = T, message = FALSE, warning = FALSE}
  export.rows %>%
      filter(`Descripción` != categorias[1]) %>%
      group_by(`Descripción`) %>%
      summarise(Total = sum(USD)) %>%
      arrange(desc(Total)) %>%
      print(n = Inf)

  export.rows %>%
      filter(`Descripción` != categorias[1]) %>%
      group_by(`Descripción`) %>%
      summarise(Total = sum(USD)) %>%
      ggplot() +
      geom_bar(aes(y = Total,
                   x = reorder(`Descripción`, Total, FUN = abs),
                   fill = Total),
               stat = 'identity') +
      coord_flip()
```

Finally, total exportations per year:

```{r, include = T, message = FALSE, warning = FALSE}
  export.cols %>%
      group_by(year) %>%
      summarise(`total export` = sum(!!sym(categorias[1]))) %>%
      print(n = Inf)

   ## Visualization
  export.rows %>%
      filter(`Descripción` != categorias[1]) %>%
      group_by(year) %>%
      summarise(Total = sum(USD)) %>%
      ggplot(aes(x = year, y = Total)) +
      geom_line() +
      geom_point() 

   ## Per state
  export.rows %>%
      filter(`Descripción` != categorias[1]) %>%
      group_by(year, state) %>%
      summarise(Total = sum(USD)) %>%
      ggplot(aes(x = year, y = Total)) +
      geom_line(aes(colour = state))+
      geom_point(aes(colour = state))
```
